Fused regression for multi-source gene regulatory network inference

### 2016-01-16 paper1.tex
hi

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become more attainable \cite{bonneau_predictive_2007, ciofani_validated_2012, carro_transcriptional_2010}. Large-scale data collection efforts have contributed to the development of high quality networks, but there remain many biological processes and organisms for which there does not exist adequate data. Furthermore, as new technologies are developed and some old ones are replaced, such as RNAseq and microarray, it becomes important to be able to combine data from multiple platforms, lest we lose valuable information from existing studies. The problem of inferring related -- but not necessarily identical -- structure from related -- but not identical -- data is ubiquitous in biology. Multi-source network inference has applications for learning multiple networks in multiple related species, for learning networks associated with distinct processes within the same species, and for learning networks based on heterogenous data sources.

Moreover, as it becomes increasingly possible to learn genome-wide regulatory networks, we can begin to compare and to test whether there is conservation of networks across species and biological processes. Our use of model organisms to study biological processes and diseases relevant to humans relies on the assumption of conservation; yet this has not been tested at the genome scale.

We present two methods for network inference based on linear estimates of gene expression dynamics, extending existing linear dynamical-systems methods for network inference \cite{bonneau_predictive_2007, arrieta-ortiz_experimentally_2015}. The core of both methods is the observation that biological information about the relatedness of genes can be used to constrain which network coefficients should be similar to one another in a multi-source network inference problem (ie orthologous TFs should regulate orthologous genes), and that these constraints can be efficiently represented as penalties in a least-squares regression problem. Our first method -- fused ridge -- uses an L2 penalty on the differences between a priori similar interactions, and is useful where the relationships between networks (similarity of genes between data sets) is reliable. In the case where both networks contain an identical set of genes and TFs, this approach can be thought of as parametrically interpolating between treating the data sources separately and combining them together. However, our method allows useful pooling of data even when the overlap between genes is incomplete. Our second method -- adaptive fusion -- uses a non-concave saturating fusion penalty to simultaneously infer the constrained networks and to learn which constraints should be relaxed (ie which parts of the network are genuinely different). With this approach, we seek to test the hypothesis of conservation of interactions across networks.

In the case of multiple species, there has been a great deal of work showing that functional conservation exists in gene regulatory networks even across large evolutionary distance \cite{satou2006gene, hinman2009evolution,tanay2005conservation,erwin2009evolution}. In our fused L2 approach, we make the assumption that if a pair of TFs is orthologous to one another, then their regulatory effects on a pair of orthologous genes should be similar. These orthology relationships form the basis of a set of constraints which favor -- but do not require -- networks in which orthologous transctiption factors regulate orthologous genes. As a result, data in one species can improve network inference performance in another species  (and vice versa). This general framework for multi-species network inference can be extended to an arbitrary number of distinct organisms, each contributing data with only partially overlapping sets of genes. This is an advantage over existing approaches to multi-species network inference, which infer only a sub-network for which orthologs exist in every species \cite{joshi_multi-species_2015}. This approach of introducting constraints on the similarity between specific regulatory interactions can be extended beyond the case of multi-species network-inference from orthology; any biological prior on similarity can be used in place of orthology. For example, we can introduce constraints that favor genes in the same operon or having similar binding sites, having similar regulators.

@@

Understanding which interactions are functionally dissimilar is of direct biological interest, but existing approaches to network inference are unable to effectively rule out conservation. Observing a large difference in the weights of regulatory interactions obtained though independent inference of multiple networks is the best evidence against conservation available with existing network inference techniques. This evidence is very weak, however; because network inference is typically underconstrainted \cite{marbach_revealing_2010-1}, there could be a different set of networks for which conservation does hold, and which fit the data almost as well. Our adaptive fusion method allows us to learn differences in networks predicted to be similar based on orthology or gene matching: because the constraints we impose favor weight configurations in which fused interaction weights are similar to one another, a solution in which a priori similar regulatory interactions are found to be different is strong evidence that they really are different.

We approach this problem by introducing a saturating penalty function based on work in statistics to develop unbiased regularization penalties for regression \cite{zhang2010nearly, fan2001variable}. The main difference between the L2 fusion approach and this adaptive fusion approach occurs when the difference between presumed analogous interactions is large despite the fusion penalty: we assume that large differences represent evidence that gene functions have diverged and represent this, in adaptive fusion, with a relaxation of the fusion penalty. The resulting cost function is non-convex and difficult to optimize; however, we can approximate its solution and obtain deeper insights into functional similarity than is available through orthology or the comparison of separately learned networks.

@@

\subsection{\textit{B. subtilis} and \textit{B. anthracis} microarray data}
We used a dataset collected for PY79, a derivative of strain 168, available on GEO with accession number GSE67023, and a dataset using BSB1, another derivative of strain 168, available at GEO with accession number GSE27219. We used two datasets for \texti{B. anthracis}, transcription profiling during iron starvation (E-MEXP-2272 on ArrayExpress), and time series over the life cycle (E-MEXP-788 on ArayExpress).

\subsection{Orthology mapping using Inparanoid}
We ran Inparanoid to obtain orthology mapping for \textit{B. subtilis} and \textit{B. anthracis} \cite{ostlund_inparanoid_2010}

@@

\noindent The matrix $\beta$ can be solved using linear regression. Because there are typically far fewer conditions than possible regressors (TFs), we introduce a ridge regularization constraint with weight $\lambda_R$ and solve
\begin{equation}
\argmin_\beta\vert \vert X\beta - Y_2 \vert \vert ^2 + \lambda_R \vert \vert \beta \vert \vert ^2
\end{equation}
This is similar to the formulation used in the Inferelator algorithm, which we extend to the case of simultaneously inferring multiple networks. 
\subsection{Learning transcription factor activity and using TFA as regressors}

@@

\subsection{Statistical approach}
Most network inference problems have more predictors than experimental samples ($p >> N$). Many methods have been proposed for penalized regression, such as ridge regression \cite{hoerl_ridge_1970}, and applied to genomic data \cite{waldron_optimized_2011, li_network-constrained_2008}. In our application, we consider the case where related outputs (genes) are likely to be affected by common inputs (transcription factors). Approaches using fusion combine statistical strength across the multiple regression tasks by introducing regularization functions reflecting prior knowledge on output structure representing how outputs are related \cite{kim_tree-guided_2012,Land1997,Chen2010,Petry2011,Hebiri2011}, and approaches which attempt to simultaneously learn the regression coefficients and the output structure \cite{sohn_joint_2012}. Fusion has been proposed as a method for utilizing correlation information from data by penalizing pairwise differences of coefficients \cite{Daye2009,Price2014}. Price et al. and Bilgrau et al. use a fused ridge estimator for jointly estimating multiple inverse covariance matrices \cite{Price2014a,Bilgrau2015}; we take a similar approach, adding an fused L2 penalty to our ridge regression to incorporate prior knowledge about relationships between input-output pairs. We also apply a penalty function bounded by a constant to produce unbiased estimators for large coefficients, combined with an L2 penalty, similar to SCAD-L2 \cite{Zeng2012}. This approach allows us to simultaneously learn the regression coefficients and evaluate the validity of this prior information.

>>>>

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become attainable \cite{bonneau_predictive_2007, ciofani_validated_2012, carro_transcriptional_2010}. Large-scale data collection efforts have contributed to the development of high quality networks which accurately recapitulate biological processes, but there remain many processes and organisms for which there does not exist adequate data. Furthermore, as new technologies are developed and some old ones are replaced, such as RNAseq and microarray, it becomes important to be able to combine data from multiple platforms, lest we lose valuable information from existing studies. The problem of inferring related -- but not necessarily identical -- structure from related -- but not identical -- data is ubiquitous in biology. Multi-source network inference has applications for learning multiple networks in multiple related species, for learning networks associated with distinct processes within the same species, and for learning networks based on heterogenous data sources.

Moreover, as it becomes possible to learn genome-wide regulatory networks, we can begin to compare and to test whether there is conservation of networks across species and biological processes. Our use of model organisms to study biological processes and diseases relevant to humans relies on the assumption of conservation; yet this has not been tested at the genome scale.

We present two methods for network inference based on linear estimates of gene expression dynamics, extending existing linear dynamical-systems methods for network inference \cite{bonneau_predictive_2007, arrieta-ortiz_experimentally_2015}. The core of both methods is the observation that biological information about the relatedness of genes can be used to constrain which network coefficients should be similar to one another in a multi-source network inference problem (ie orthologous TFs should regulate orthologous genes), and that these constraints can be efficiently represented as penalties in a least-squares regression problem. Our approach uses predictions of similarity of putative interactions, to improve our ability to accurately describe TF-gene relationships on a genome-wide scale.

Our first method -- fused ridge -- uses an L2 penalty on the differences between a priori similar interactions, and is useful where the relationships between networks (similarity of genes between data sets) is reliable. In the case where both networks contain an identical set of genes and TFs, this approach can be thought of as parametrically interpolating between treating the data sources separately and combining them together. However, our method allows useful pooling of data even when the overlap between genes is incomplete. Our second method -- adaptive fusion -- uses a non-concave saturating fusion penalty to simultaneously infer the constrained networks and to learn which constraints should be relaxed (ie which parts of the network are genuinely different). With this approach, we seek to test the hypothesis of conservation of interactions across networks.

In the case of multiple species, there has been a great deal of work showing that functional conservation exists in gene regulatory networks even across large evolutionary distance \cite{satou2006gene, hinman2009evolution,tanay2005conservation,erwin2009evolution}. In our fused L2 approach, we make the assumption that for closely related species, if a pair of TFs is orthologous to one another, then their regulatory effects on a pair of orthologous genes should be similar. These orthology relationships form the basis of a set of constraints which favor -- but do not require -- networks in which orthologous transctiption factors regulate orthologous genes. As a result, data in one species can improve network inference performance in another species  (and vice versa). This general framework for multi-species network inference can be extended to an arbitrary number of distinct organisms, each contributing data with only partially overlapping sets of genes. This is an advantage over existing approaches to multi-species network inference, which infer only a sub-network for which orthologs exist in every species \cite{joshi_multi-species_2015}. This approach of introducting constraints on the similarity between specific regulatory interactions can be extended beyond the case of multi-species network-inference from orthology; any biological prior on similarity can be used in place of orthology. For example, we can introduce constraints that favor genes in the same operon or having similar binding sites, having similar regulators.

@@

Understanding which interactions are functionally dissimilar is of direct biological interest, but existing approaches to network inference are unable to effectively rule out conservation. Observing a large difference in the weights of regulatory interactions obtained though independent inference of multiple networks is the best evidence against conservation available with existing network inference techniques. This evidence is very weak, however; because network inference is typically underconstrainted \cite{marbach_revealing_2010-1}, there could be a different set of networks for which conservation does hold, and which fit the data almost as well. We propose, with our adaptive fusion, an approach where network inference occurs simultaneously with the evaluation of conservation, to gather stronger evidence of divergence.

We approach this problem by introducing a saturating penalty function based on work in statistics to develop unbiased regularization penalties for regression \cite{zhang2010nearly, fan2001variable}. The main difference between the L2 fusion approach and this adaptive fusion approach occurs when the difference between presumed analogous interactions is large despite the fusion penalty: we assume that large differences represent evidence that gene functions have diverged and represent this, in adaptive fusion, with a relaxation of the fusion penalty. The resulting cost function is non-convex and difficult to optimize; however, we can approximate its solution and obtain deeper insight into functional similarity than is available through orthology or the comparison of separately learned networks.

@@

\subsection{\textit{B. subtilis} and \textit{B. anthracis} data and orthology}
We used a dataset collected for PY79, a derivative of strain 168, available on GEO with accession number GSE67023, and a dataset using BSB1, another derivative of strain 168, available at GEO with accession number GSE27219. We used two datasets for \texti{B. anthracis}, transcription profiling during iron starvation (E-MEXP-2272 on ArrayExpress), and time series over the life cycle (E-MEXP-788 on ArayExpress). We ran Inparanoid to obtain orthology mapping for \textit{B. subtilis} and \textit{B. anthracis} \cite{ostlund_inparanoid_2010}

@@

\noindent The matrix $\beta$ can be solved using linear regression. Because there are typically far fewer conditions than possible regressors (TFs), we introduce a ridge regularization constraint with weight $\lambda_R$ and solve
\begin{equation}
\argmin_\beta\vert \vert X\beta - Y_2 \vert \vert ^2 + \lambda_R \vert \vert \beta \vert \vert ^2
\end{equation}
This is similar to the formulation used in the Inferelator algorithm, which we extend to the case of simultaneously inferring multiple networks. 
\subsection{Estimating transcription factor activity and using TFA as regressors}
When there exists a set of prior known interactions, we are able to estimate transcription factor activity (TFA) using network component analysis \cite{liao2003network}, as in \cite{arrieta-ortiz_experimentally_2015, fu_reconstructing_2011}, and use TFA as explanatory variables instead of transcription factor expression.

@@

\subsection{Statistical approach and background}
Most network inference problems have more predictors than experimental samples ($p >> N$). Many methods have been proposed for penalized regression, such as ridge regression \cite{hoerl_ridge_1970}, and applied to genomic data \cite{waldron_optimized_2011, li_network-constrained_2008}. In our application, we consider the case where related outputs (genes) are likely to be affected by common inputs (transcription factors). Fusion has been proposed as a method for utilizing correlation information from data by penalizing pairwise differences of coefficients \cite{Daye2009,Price2014}. Approaches using fusion combine statistical strength across the multiple regression tasks by introducing regularization functions reflecting prior knowledge on output structure representing how outputs are related \cite{kim_tree-guided_2012,Land1997,Chen2010,Petry2011,Hebiri2011}. Price et al. and Bilgrau et al. use a fused ridge estimator for jointly estimating multiple inverse covariance matrices \cite{Price2014a,Bilgrau2015}; we take a similar approach, adding an fused L2 penalty to our ridge regression to incorporate prior knowledge about relationships between input-output pairs. Sohn et al. attempt to simultaneously learn the regression coefficients and the output structure \cite{sohn_joint_2012}. We also take a simultaneous approach, where we apply a penalty function bounded by a constant to produce unbiased estimators for large coefficients, combined with an L2 penalty, similar to SCAD-L2 \cite{Zeng2012}. This approach allows us to simultaneously learn the regression coefficients and evaluate the validity of our prior information.
